---
output:
  md_document:
    variant: markdown_github
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.path = "README/README-",
  message = FALSE, warning = FALSE, fig.asp = 1, fig.align = 'center'
)
```

sphunif
=======

```{r, echo = FALSE, results = 'asis'}
cat(
  badger::badge_license(license = "GPLv3", color = "blue",
                        url = "https://www.gnu.org/licenses/gpl-3.0"),
  badger::badge_travis(ref = NULL),
	badger::badge_cran_release(color = "green"),
	badger::badge_cran_download(pkg = NULL, type = "grand-total"),
	badger::badge_cran_download(pkg = NULL, type = "last-month")
)
```

<!-- <img src="" alt="sphunif  hexlogo" align="right" width="200" style="padding: 0 15px; float: right;"/> -->


## Overview

Implementation of circa 30 tests of uniformity on the circle, sphere, and hypersphere. Software companion for the (evolving) review "*An overview of uniformity tests on the hypersphere*" (García-Portugués and Verdebout, 2018) and the paper "*On a projection-based class of uniformity tests on the hypersphere*" (García-Portugués, Navarro-Esteban and Cuesta-Albertos, 2020).

## Installation

Get the latest version from GitHub:

```{r, install-devtools, eval = FALSE}
# Install the package
library(devtools)
install_github("egarpor/sphunif")

# Load package
library(sphunif)
```

```{r, load, echo = FALSE}
# Load package
library(sphunif)
```

## Usage

### Circular data

You want to test if a sample of *circular* data is uniformly distributed. For example, the following circular uniform sample in *radians*:

```{r, cir_data}
set.seed(987202226)
cir_data <- runif(n = 10, min = 0, max = 2 * pi)
```

Call the main function in the `sphunif` package, `unif_test`, specifying the `type` of test to be performed. For example, the `"Kuiper"` test:

```{r, unif_test_cir}
library(sphunif)
unif_test(data = cir_data, type = "Kuiper", verbose = FALSE) # An htest object
```

By default, the calibration of the test statistic is done by Monte Carlo. This can be changed with `p_value = "asymp"` to employ asymptotic distributions (faster, but not available for all tests):

```{r, asymp}
unif_test(data = cir_data, type = "Kuiper", p_value = "MC",
          verbose = FALSE) # Monte Carlo
unif_test(data = cir_data, type = "Kuiper", p_value = "asymp") # Asymp. distr.
```

You can perform *several* tests with a *single* call to `unif_test`. Choose the available circular uniformity tests from

```{r, avail_cir}
avail_cir_tests
```

For example:

```{r, unif_test_avail_cir}
unif_test(data = cir_data, type = c("Kuiper", "Watson", "Ajne"),
          verbose = FALSE) # A *list* of htest objects
```

### Spherical data

You want to test if a sample of *spherical* data is uniformly distributed. Consider a *non*-uniformly-generated sample of directions in **Cartesian coordinates**:

```{r, sph_data}
# Sample data on S^2
set.seed(987202226)
theta <- runif(n = 50, min = 0, max = 2 * pi)
phi <- runif(n = 50, min = 0, max = pi)
sph_data <- cbind(cos(theta) * sin(phi), sin(theta) * sin(phi), cos(phi))
```

The available spherical uniformity tests:

```{r, avail_sph}
avail_sph_tests
```

The default `type = "all"`  equals `type = avail_sph_tests`:

```{r, unif_test_avail_sph}
unif_test(data = sph_data, type = "all", p_value = "MC", verbose = FALSE)
unif_test(data = sph_data, type = "Rayleigh", p_value = "asymp")
```

### Higher dimensions

The *hyperspherical* setting is treated analogously to the spherical setting, and the available tests are exactly the same (`avail_sph_tests`). An example of testing uniformity with a sample of size `100` on the $9$-sphere:

```{r, hyp_data}
# Sample data on S^9
set.seed(987202226)
hyp_data <- r_unif_sph(n = 50, p = 10)

# Test
unif_test(data = hyp_data, type = "Rayleigh", p_value = "asymp")
```

## Data applications in astronomy

The data applications in García-Portugués, Navarro-Esteban and Cuesta-Albertos (2020) can
be reproduced through the script [data-application-ecdf.R](https://github.com/egarpor/sphunif/blob/master/application/data-application-ecdf.R). The code snippet below illustrates the testing of the uniformity of orbits of planets and long-period comets.

```{r, sunspots_births}
## Planets

# Load data
data("planets")

# Add normal vectors
planets$normal <- cbind(sin(planets$i) * sin(planets$om),
                       -sin(planets$i) * cos(planets$om),
                       cos(planets$i))

# Tests without Pluto
unif_test(data = planets$normal[-9, ], type = c("PCvM", "PAD", "PRt"),
          p_value = "MC")

## Comets

# Load data
data("comets")

# Add normal vectors
comets$normal <- cbind(sin(comets$i) * sin(comets$om),
                       -sin(comets$i) * cos(comets$om),
                       cos(comets$i))

# Exclude the C/1882 R1-X (Great September comet) records with X = B, C, D
comets_ccf2009 <- comets[comets$ccf2009, ][-c(13:15), ]

# Tests for the data in Cuesta-Albertos et al. (2009)
tests_ccf2009 <- unif_test(data = comets_ccf2009$normal,
                           type = c("PCvM", "PAD", "PRt"),
                           p_value = "asymp", K_max = 5e4)
tests_ccf2009
```

## References

García-Portugués, E., Navarro-Esteban, P., and Cuesta-Albertos, J. A. (2020). On a projection-based class of uniformity tests on the hypersphere. *arXiv:2008.09897*. <https://arxiv.org/abs/2008.09897>

García-Portugués, E. and Verdebout, T. (2018). An overview of uniformity tests
on the hypersphere. *arXiv:1804.00286*. <https://arxiv.org/abs/1804.00286>
