---
title: "Non-uniform distributions in the `sphunif` package"
author: "Eduardo García-Portugués"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    toc: yes
    fig_width: 7
    fig_height: 7
vignette: >
  %\VignetteIndexEntry{Non-uniform distributions in the `sphunif` package}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  fig.asp = 1
)
```

## Common settings

The workhorse functions for sampling non-uniform data are:

* `r_alt`. Simulates the following distributions: von Mises--Fisher (vMF), a specific instance of a Mixture of vMF (MvMF), Angular Central Gaussian (ACG), Small Circle (SC), and Watson (W). The departure from uniformity is controlled by $\kappa\ge 0$ ($\kappa=0$ gives uniformity).
* `r_locdev`. Serves to simulate local projected alternatives, specially if employed together with `cutoff_locdev` for the following Projected distributions: Rothman (PRt), Cramér--von Mises (PCvM), and Anderson--Darling (PAD). The departure from uniformity is controlled by $0\le\kappa\le 1$ ($\kappa=0$ gives uniformity).

The following sections illustrate the use of these functions for the next settings.

```{r, sphunif}
# Load packages
library(sphunif)
library(scatterplot3d)

# Sample size
n <- 2e2

# kappas
kappas <- c(0.25, 1.00)

# Seed
seed <- 987202226

# Plotting function for circular sample
plot_cir <- function(samp, kappa) {

  plot(runif(nrow(samp), 0.85, 1.15) * samp, xlab = "x1", ylab = "x2",
       pch = 19, cex = 0.25, xlim = c(-1.15, 1.15), ylim = c(-1.15, 1.15),
       main = paste("kappa =", kappa))

}

# Plotting function for spherical sample
plot_sph <- function(samp, kappa) {

  scatterplot3d(samp, xlab = "x1", ylab = "x2", zlab = "x3",
                xlim = c(-1, 1), ylim = c(-1, 1), zlim = c(-1, 1),
                pch = 19, cex.symbols = 0.25, main = paste("kappa =", kappa))

}
```

## `r_alt`

### Von Mises--Fisher (vMF)

#### $p=2$

```{r, vMF-2}
set.seed(seed)
par(mfrow = c(1, 2))
for (k in kappas) {

  plot_cir(r_alt(n = n, p = 2, scenario = "vMF", kappa = k)[, , 1], kappa = k)

}
```

#### $p=3$

```{r, vMF-3}
set.seed(seed)
par(mfrow = c(1, 2))
for (k in kappas) {

  plot_sph(r_alt(n = n, p = 3, scenario = "vMF", kappa = k)[, , 1], kappa = k)

}
```

### Mixture of von Mises--Fisher (MvMF)

#### $p=2$

```{r, MvMF-2}
set.seed(seed)
par(mfrow = c(1, 2))
for (k in (16 * kappas)) {

  plot_cir(r_alt(n = n, p = 2, scenario = "MvMF", kappa = k)[, , 1], kappa = k)

}
```

#### $p=3$

```{r, MvMF-3}
set.seed(seed)
par(mfrow = c(1, 2))
for (k in (16 * kappas)) {

  plot_sph(r_alt(n = n, p = 3, scenario = "MvMF", kappa = k)[, , 1], kappa = k)

}
```

### Angular Central Gaussian (ACG)

#### $p=2$

```{r, ACG-2}
set.seed(seed)
par(mfrow = c(1, 2))
for (k in (1.5 * kappas)) {

  plot_cir(r_alt(n = n, p = 2, scenario = "ACG", kappa = k)[, , 1], kappa = k)

}
```

#### $p=3$

```{r, ACG-3}
set.seed(seed)
par(mfrow = c(1, 2))
for (k in (1.5 * kappas)) {

  plot_sph(r_alt(n = n, p = 3, scenario = "ACG", kappa = k)[, , 1], kappa = k)

}
```

### Watson (W)

#### $p=2$

```{r, W-2}
set.seed(seed)
par(mfrow = c(1, 2))
for (k in kappas) {

  plot_cir(r_alt(n = n, p = 2, scenario = "W", kappa = k)[, , 1], kappa = k)

}
```

#### $p=3$

```{r, W-3}
set.seed(seed)
par(mfrow = c(1, 2))
for (k in kappas) {

  plot_sph(r_alt(n = n, p = 3, scenario = "W", kappa = k)[, , 1], kappa = k)

}
```

### Small Circle (SC)

#### $p=2$

```{r, SC-2}
set.seed(seed)
par(mfrow = c(1, 2))
for (k in kappas) {

  plot_cir(r_alt(n = n, p = 2, scenario = "SC", kappa = k)[, , 1], kappa = k)

}
```

#### $p=3$

```{r, SC-3}
set.seed(seed)
par(mfrow = c(1, 2))
for (k in kappas) {

  plot_sph(r_alt(n = n, p = 3, scenario = "SC", kappa = k)[, , 1], kappa = k)

}
```

## `r_locdev`

### Projected Rothman (PRt)

#### $p=2$

```{r, PRt-2}
set.seed(seed)
par(mfrow = c(1, 2))
Rt_t <- 1 / 3
for (k in kappas) {

  f <- f_locdev_Pn(p = 2, type = "PRt", Rothman_t = Rt_t)
  F_inv <- F_inv_from_f(f = f, p = 2)
  plot_cir(r_locdev(n = n, mu = c(0, 1), F_inv = F_inv, kappa = k), kappa = k)

}
```

#### $p=3$

```{r, PRt-3}
set.seed(seed)
par(mfrow = c(1, 2))
Rt_t <- 1 / 3
for (k in kappas) {

  f <- f_locdev_Pn(p = 3, type = "PRt", Rothman_t = Rt_t)
  F_inv <- F_inv_from_f(f = f, p = 3)
  plot_sph(r_locdev(n = n, mu = c(0, 0, 1), F_inv = F_inv, kappa = k),
           kappa = k)

}
```

### Projected Cramér--von Mises (PCvM)

#### $p=2$

```{r, PCvM-2}
set.seed(seed)
par(mfrow = c(1, 2))
for (k in kappas) {

  f <- f_locdev_Pn(p = 2, type = "PCvM")
  F_inv <- F_inv_from_f(f = f, p = 2)
  plot_cir(r_locdev(n = n, mu = c(0, 1), F_inv = F_inv, kappa = k), kappa = k)

}
```

#### $p=3$

```{r, PCvM-3}
set.seed(seed)
par(mfrow = c(1, 2))
for (k in kappas) {

  f <- f_locdev_Pn(p = 3, type = "PCvM")
  F_inv <- F_inv_from_f(f = f, p = 3)
  plot_sph(r_locdev(n = n, mu = c(0, 0, 1), F_inv = F_inv, kappa = k),
           kappa = k)

}
```

### Projected Anderson--Darling (PAD)

#### $p=2$

```{r, PAD-2}
set.seed(seed)
par(mfrow = c(1, 2))
for (k in kappas) {

  f <- f_locdev_Pn(p = 2, type = "PAD")
  F_inv <- F_inv_from_f(f = f, p = 2)
  plot_cir(r_locdev(n = n, mu = c(0, 1), F_inv = F_inv, kappa = k), kappa = k)

}
```

#### $p=3$

```{r, PAD-3}
set.seed(seed)
par(mfrow = c(1, 2))
for (k in kappas) {

  f <- f_locdev_Pn(p = 3, type = "PCvM")
  F_inv <- F_inv_from_f(f = f, p = 3)
  plot_sph(r_locdev(n = n, mu = c(0, 0, 1), F_inv = F_inv, kappa = k),
           kappa = k)

}
```