---
title: "`sphunif`: Uniformity Tests on the Circle, Sphere, and Hypersphere"
author: "Eduardo García-Portugués and Thomas Verdebout"
date: "`r Sys.Date()`, v`r packageVersion('sphunif')`"
bibliography: sphunif.bib
biblio-style: apalike-custom
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{sphunif: Uniformity Tests on the Circle, Sphere, and Hypersphere}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Just give me a quick example!

### Circular data

Suppose that you want to test if a sample of *circular* data is uniformly distributed. For example, the following circular uniform sample in **radians**:

```{r, cir_data}
set.seed(987202226)
cir_data <- runif(n = 30, min = 0, max = 2 * pi)
```

Then you can call the main function in the `sphunif` package, `unif_test`, specifying the `type` of test to be performed. For example, the Watson (omnibus) test:

```{r, unif_test_cir}
library(sphunif)
unif_test(data = cir_data, type = "Watson", verbose = FALSE) # An htest object
```

By default, the calibration of the test statistic is done by Monte Carlo. This can be changed with `p_value = "asymp"` to use the asymptotic distribution:

```{r, asymp}
unif_test(data = cir_data, type = "Watson", p_value = "MC",
          verbose = FALSE) # Monte Carlo
unif_test(data = cir_data, type = "Watson", p_value = "asymp") # Asymp. distr.
```

You can perform *several* tests within a *single* call to `unif_test`. Choose the available circular uniformity tests from

```{r, avail_cir}
avail_cir_tests
```

For example, you can use the Projected Anderson--Darling (@Garcia-Portugues2020b, also an omnibus test) test and the Watson test:

```{r, unif_test_avail_cir}
unif_test(data = cir_data, type = c("PAD", "Watson"),
          verbose = FALSE) # A *list* of htest objects
```

@Garcia-Portugues2020a gives a review of different tests of uniformity on the circle. Section 5.1 in @Pewsey2021 also gives an overview of recent advances.

### Spherical data

Suppose now that you want to test if a sample of *spherical* data is uniformly distributed. Consider a *non*-uniformly-generated^[Uniformly-distributed polar coordinates do not translate into uniformly-distributed Cartesian coordinates!] sample of directions in **Cartesian coordinates**, such as:

```{r, sph_data}
# Sample data on S^2
set.seed(987202226)
theta <- runif(n = 30, min = 0, max = 2 * pi)
phi <- runif(n = 30, min = 0, max = pi)
sph_data <- cbind(cos(theta) * sin(phi), sin(theta) * sin(phi), cos(phi))
```

The available spherical uniformity tests:

```{r, avail_sph}
avail_sph_tests
```

See again @Garcia-Portugues2020a for a review of tests of uniformity on the sphere and complementary Section 5.1 in @Pewsey2021.

The default `type = "all"` equals `type = avail_sph_tests`, which might be too much for standard analysis:

```{r, unif_test_avail_sph}
unif_test(data = sph_data, type = "all", p_value = "MC", M = 1e3,
          verbose = FALSE)
unif_test(data = sph_data, type = "Rayleigh", p_value = "asymp")
```

### Higher dimensions

The *hyperspherical* setting is treated analogously to the spherical setting, and the available tests are exactly the same (`avail_sph_tests`). Here is an example of testing uniformity with a sample of size `100` on the $9$-sphere:

```{r, hyp_data}
# Sample data on S^9
hyp_data <- r_unif_sph(n = 30, p = 10)

# Test
unif_test(data = hyp_data, type = "Rayleigh", p_value = "asymp")
```


## A data example: are Venusian craters uniformly distributed?

The following snippet partially reproduces the data application in @Garcia-Portugues2021 on testing the uniformity of known Venusian craters:

```{r, venus}
# Load spherical data
data(venus)
head(venus)
nrow(venus)

# Compute Cartesian coordinates from polar coordinates
venus$X <- cbind(cos(venus$theta) * cos(venus$phi),
                 sin(venus$theta) * cos(venus$phi),
                 sin(venus$phi))

# Test uniformity using the Projected Cramér-von Mises and Projected
# Anderson-Darling tests on S^2, both using asymptotic distributions
unif_test(data = venus$X, type = c("PCvM", "PAD"), p_value = "asymp")
```

## Simulation studies done simple

`unif_stat` allows to compute several statistics to different samples within a single call, thus facilitating Monte Carlo experiments:

```{r, unif_stat_vec}
# M samples of size n on S^2
samps_sph <- r_unif_sph(n = 30, p = 3, M = 10)

# Apply all statistics to the M samples
unif_stat(data = samps_sph, type = "all")
```

Additionally, `unif_stat_MC` is an utility for performing the previous simulation through a convenient parallel wrapper:

```{r, MC}
# Break the simulation in 10 chunks of tasks to be divided between 2 cores
sim <- unif_stat_MC(n = 30, type = "all", p = 3, M = 1e4, cores = 2,
                    chunks = 10, verbose = FALSE)

# Critical values for test statistics
sim$crit_val_MC

# Test statistics
head(sim$stats_MC)

# Power computation using a pre-built sampler for the alternative
pow <- unif_stat_MC(n = 30, type = "all", p = 3, M = 1e4, cores = 2,
                    chunks = 10, r_H1 = r_alt, crit_val = sim$crit_val_MC,
                    alt = "vMF", kappa = 1, verbose = FALSE)
pow$power_MC

# How to use a custom sampler according to ?unif_stat_MC
r_H1 <- function(n, p, M, l = 1) {

  samp <- array(dim = c(n, p, M))
  for (j in 1:M) {

    samp[, , j] <- mvtnorm::rmvnorm(n = n, mean = c(l, rep(0, p - 1)),
                                    sigma = diag(rep(1, p)))
    samp[, , j] <- samp[, , j] / sqrt(rowSums(samp[, , j]^2))

  }
  return(samp)

}
pow <- unif_stat_MC(n = 30, type = "all", p = 3, M = 1e2, cores = 2,
                    chunks = 5, r_H1 = r_H1, crit_val = sim$crit_val_MC,
                    verbose = FALSE)
pow$power_MC
```

`unif_stat_MC` can be used for constructing the Monte Carlo calibration necessary for `unif_test`, either for providing a rejection rule based on `$crit_val_MC` or for approximating the $p$-value by `$stats_MC`.

```{r, crit_val}
# Using precomputed critical values
ht1 <- unif_test(data = samps_sph[, , 1], type = "Rayleigh",
                 p_value = "crit_val", crit_val = sim$crit_val_MC,
                 verbose = FALSE)
ht1
ht1$reject

# Using precomputed Monte Carlo statistics
ht2 <- unif_test(data = samps_sph[, , 1], type = "Rayleigh",
                 p_value = "MC", stats_MC = sim$stats_MC, verbose = FALSE)
ht2
ht2$reject

# Faster than
unif_test(data = samps_sph[, , 1], type = "Rayleigh", p_value = "MC",
          verbose = FALSE)
```

## References

